---
title: "EBOV sequence datasets"
author: "Louis du Plessis"
date: \today
fontsize: 10pt
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    keep_tex: false
    fig_crop: false
---

# Summary

Get a few smaller subsets of the dataset on https://github.com/ebov/space-time.


```{r rsetup, echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.path="figs/", fig.width=8, fig.height=4}

    rm(list = ls())
    library(gplots)
    library(bdskytools)
    library(knitr)
    
    source("../scripts/PlotUtils.R")

    addStats <- function(stats, dataset, name="data") {
      
        stats$names     <- c(stats$names,     name)
        stats$sequences <- c(stats$sequences, nrow(dataset))
        stats$earliest  <- c(stats$earliest,  min(dataset$date))
        stats$latest    <- c(stats$latest,    max(dataset$date))
        stats$range     <- c(stats$range,     diff(range(dataset$date)))

        return(stats)  
    }
    
    
    #' Quick function for downsampling to some fraction, approximately stratified 
    #' by month, with a minimum monthly count cutoff
    #' 
    #' Returns the indices in the original vector to sample
    downsampleStratifiedByMonth <- function(sampledates, frac, cutoff=5) {
    
      # Get monthly counts
      monthlycounts <- hist(sampledates, breaks='months', freq=TRUE, plot = FALSE)
      
      # Get new monthly counts after downsampling
      monthlysamples <- c()
      for (m in monthlycounts$counts) {
          count <- ceiling(frac*m)
          if (count < cutoff) {
              monthlysamples <- c(monthlysamples, m)
          } else {
              monthlysamples <- c(monthlysamples, count)
          }
      }
      
      
      # Get indices for each bin in the histogram and subsample
      # By default histogram cells are right closed, left open intervals
      breakdates <- as.Date(monthlycounts$breaks, origin=dateorigin)
      indices    <- c()
      for (i in 1:(length(breakdates)-1)) {
          samples <- which(sampledates > breakdates[i] & sampledates <= breakdates[i+1])
          # print(paste(length(samples), monthlycounts$counts[i]))
          indices <- c(indices, sample(samples, monthlysamples[i], replace=FALSE))
      }
      return(indices)
      
    }

    
    inputpath  <- "../data/sequence/Data/"
    outputpath <- "../results/datasets/"
    
    # Preprocessing
    seqdb <- read.csv(paste0(inputpath, "Makona_1610_metadata_2016-06-23.csv"), row.names = 1)
    
    # Column names to rename
    renamemap <- list("final_label"  ="label",
                      "prop_ambig."  ="proportion_ambiguous",
                      "imputed_date."="imputed_date")
    
    # Rename columns, add NAs and remove commas
    cols <- names(seqdb)
    for (i in 1:length(cols)) {
      print(cols[i])
      seqdb[[cols[i]]][which(seqdb[cols[i]] == "")]  <- NA
      seqdb[[cols[i]]][which(seqdb[cols[i]] == "?")] <- NA
      
      collevels <- levels(seqdb[[cols[i]]])
      commas <- grep(",", collevels)
      for (c in commas) {
          collevels[c] <- gsub(",","",collevels[c])
      }
      levels(seqdb[[cols[i]]]) <- collevels
      
      if (cols[i] %in% names(renamemap))
          cols[i] <- renamemap[[cols[i]]]
    }
    names(seqdb) <- cols
    seqdb <- droplevels(seqdb)
    
    # Convert to dates
    seqdb$date            <- as.Date(seqdb$date, format="%d/%m/%Y")
    seqdb$imputed_date    <- as.Date(seqdb$imputed_date, format="%d/%m/%Y")
    seqdb$collection_date <- as.Date(seqdb$collection_date, format="%d/%m/%Y")
    
    # Save file
    write.csv(seqdb, paste0(outputpath,"complete.csv"), quote=FALSE, row.names=FALSE)
    
    # Track statistics of datasets
    stats <- list(names=c(), sequences=c(), earliest=c(), latest=c(), range=c())
    
    # Load maps
    SLE_adm2 <- readRDS("../data/maps/SLE_adm2.rds")
    GIN_adm2 <- readRDS( "../data/maps/GIN_adm2.rds")
    LBR_adm1 <- readRDS( "../data/maps/LBR_adm1.rds")
    LBR_adm1$NAME_2 <- LBR_adm1$NAME_1
    
    # Ordered countries and locations
    countries <- as.factor(c(rep("LBR",13), rep("SLE",15), rep("GIN",19)))
    locations <- c("Bomi","Bong","Gbarpolu","GrandBassa","GrandCapeMount","GrandKru","Lofa","Margibi","Montserrado","Nimba","RiverCess","RiverGee","Sinoe",
                   "Kailahun","Kenema","Kono","Bo","Bonthe","Moyamba","Pujehun","Bombali","Koinadugu","Tonkolili","Kambia","PortLoko","WesternArea","WesternRural","WesternUrban",
                   "Conakry","Boke","Fria","Coyah","Dubreka","Forecariah","Kindia","Dalaba",
                   "Faranah","Kissidougou","Kankan","Kerouane","Kouroussa","Siguiri","Beyla","Gueckedou","Lola","Macenta","Nzerekore")
    #regions   <- as.factor(c(rep("LBR",13), rep("East SLE",3),  rep("South SLE",4), rep("North SLE",5), rep("West SLE",3),rep("West GIN",8),rep("East GIN",11)))
    
```

---    
    
\clearpage

# Datasets

## Complete dataset 


```{r complete-date, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}

    # Reload 
    #seqdb <- read.csv(paste0(outputpath,"EBOV_ALL.csv"), row.names = 1)
    plotSamplingHist(seqdb$date, ymax=80, ylab="", xlab=paste0("Complete dataset (weekly counts)\n(",nrow(seqdb)," sequences)"), cex.years=1, cex.months=0.7, cex.axis=0.8)
    plotSamplingHist(seqdb$date, ymax=250, ylab="", xlab=paste0("Complete dataset (monthly counts)\n(",nrow(seqdb)," sequences)"), cex.years=1, cex.months=0.7, cex.axis=0.8, breaks='months', col=pal.dark(cgreen))
    stats <- addStats(stats, seqdb, "complete")
    
```

```{r complete-geo, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=3.5, fig.height=3}

    plotMapGradient(getDistrictCounts(seqdb), GIN_adm2, numlabels=5, lwd=0.5, col=pal.dark(cgreen))
    plotMapGradient(getDistrictCounts(seqdb), SLE_adm2, numlabels=5, lwd=0.5, col=pal.dark(cblue))
    plotMapGradient(getDistrictCounts(seqdb), LBR_adm1, numlabels=5, lwd=0.5, col=pal.dark(corange))
    
    par(mar=c(2,2,2,2)+0.1)
    counts <- table(seqdb$country)
    pie(counts, paste0(names(counts), " (", counts,")"), border="white", col=pal.dark(c(cgreen, corange, cblue)), cex=0.8)

```

---

\clearpage

## Southeast Sierra Leone

```{r region-sle-southeast-date, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}
    
    locs <- c("Kailahun", "Kenema", "Kono", "Bo", "Bonthe", "Moyamba", "Pujehun")
    SESLE <- droplevels(seqdb[which(seqdb$location %in% locs),])
    write.csv(SESLE, paste0(outputpath,"region-sle-southeast.csv"), quote=FALSE, row.names=FALSE)
    
    plotSamplingHist(SESLE$date, daterange=range(seqdb$date), ymax=50, ylab="", xlab=paste0("Southeast Sierra Leone\n(",nrow(SESLE)," sequences)"), cex.years=1, cex.months=0.7, cex.axis=0.8)
    stats <- addStats(stats, SESLE, "region-sle-southeast")
```

```{r region-sle-southeast-geo, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=3.5, fig.height=3}

    par(mar=c(5,4,0,0)+0.1)
    plotProportions(table(SESLE$location), labels=locations[which(countries == "SLE")], prop=FALSE, cex.axis=0.7, cex.names=0.7, cex.counts=0.6, col=pal.dark(cblue))
    plotMapGradient(getDistrictCounts(SESLE), SLE_adm2, numlabels=5, lwd=0.5, col=pal.dark(cblue))

```


## Western Sierra Leone

```{r region-sle-west-date, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}
    
    locs <- c("WesternArea", "WesternRural", "WesternUrban")
    WSLE <- droplevels(seqdb[which(seqdb$location %in% locs),])
    write.csv(WSLE, paste0(outputpath,"region-sle-west.csv"), quote=FALSE, row.names=FALSE)
    
    plotSamplingHist(WSLE$date, daterange=range(seqdb$date), ymax=50, ylab="", xlab=paste0("Western Sierra Leone\n(",nrow(WSLE)," sequences)"), cex.years=1, cex.months=0.7, cex.axis=0.8)
    stats <- addStats(stats, WSLE, "region-sle-west")
```

```{r region-sle-west-geo, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=3.5, fig.height=3}

    par(mar=c(5,4,0,0)+0.1)
    plotProportions(table(WSLE$location), labels=locations[which(countries == "SLE")], prop=FALSE, cex.axis=0.7, cex.names=0.7, cex.counts=0.6, col=pal.dark(cblue))
    plotMapGradient(getDistrictCounts(WSLE), SLE_adm2, numlabels=5, lwd=0.5, col=pal.dark(cblue))

```


\clearpage


## Eastern Sierra Leone

```{r region-sle-east-date, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}
    
    locs <- c("Kailahun", "Kenema", "Kono")
    ESLE <- droplevels(seqdb[which(seqdb$location %in% locs),])
    write.csv(ESLE, paste0(outputpath,"region-sle-east.csv"), quote=FALSE, row.names=FALSE)
    
    plotSamplingHist(ESLE$date, daterange=range(seqdb$date), ymax=50, ylab="", xlab=paste0("Eastern Sierra Leone\n(",nrow(ESLE)," sequences)"), cex.years=1, cex.months=0.7, cex.axis=0.8)
    stats <- addStats(stats, ESLE, "region-sle-east")
```


```{r region-sle-east-geo, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=3.5, fig.height=3}

    par(mar=c(5,4,0,0)+0.1)
    plotProportions(table(ESLE$location), labels=locations[which(countries == "SLE")], prop=FALSE, cex.axis=0.7, cex.names=0.7, cex.counts=0.6, col=pal.dark(cblue))
    plotMapGradient(getDistrictCounts(ESLE), SLE_adm2, numlabels=5, lwd=0.5, col=pal.dark(cblue))

```

\clearpage

## Liberia

```{r country-lbr-date, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}
    
    LBR <- droplevels(seqdb[which(seqdb$country == "LBR"),])
    write.csv(LBR, paste0(outputpath,"country-lbr.csv"), quote=FALSE, row.names=FALSE)
    
    plotSamplingHist(LBR$date, daterange=range(seqdb$date), ymax=80, ylab="", xlab=paste0("Liberia\n(",nrow(LBR)," sequences)"), cex.years=1, cex.months=0.7, cex.axis=0.8, col=pal.dark(cblue))
    stats <- addStats(stats, LBR, "country-lbr")

```

```{r country-lbr-geo, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=3.5, fig.height=3}

    par(mar=c(5,4,0,0)+0.1)
    plotProportions(table(LBR$location), labels=levels(LBR$location), prop=FALSE, cex.axis=0.7, cex.names=0.7, cex.counts=0.6, col=pal.dark(cblue))
    plotMapGradient(getDistrictCounts(LBR), LBR_adm1, numlabels=5, lwd=0.5, col=pal.dark(cblue))

```


\clearpage


## Early outbreak

All sequences until 8 August (WHO declares public health emergency). 

```{r special-early-date, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}

    early  <- droplevels(seqdb[which(seqdb$date < as.Date("2014-08-08")),])
    write.csv(early, paste0(outputpath,"special-early.csv"), quote=FALSE, row.names=FALSE)
    
    plotSamplingHist(early$date, daterange=range(seqdb$date), ymax=50, ylab="", xlab=paste0("Early outbreak\n(",nrow(early)," sequences)"), cex.years=1, cex.months=0.7, cex.axis=0.8, col=pal.dark(cblue))
    stats <- addStats(stats, early, "special-early")

    plotSamplingHist(early$date, daterange=range(early$date), ymax=50, ylab="", xlab=paste0("Early outbreak\n(",nrow(early)," sequences)"), cex.years=1, cex.months=0.7, cex.axis=0.8, col=pal.dark(cblue))

```

\clearpage

```{r special-early-geo, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=3.5, fig.height=3}
    plotMapGradient(getDistrictCounts(early[which(early$country == "GIN"),]), GIN_adm2, numlabels=5, lwd=0.5, col=pal.dark(cgreen))
    plotMapGradient(getDistrictCounts(early[which(early$country == "SLE"),]), SLE_adm2, numlabels=5, lwd=0.5, col=pal.dark(cblue))
    plotMapGradient(getDistrictCounts(early[which(early$country == "LBR"),]), LBR_adm1, numlabels=5, lwd=0.5, col=pal.dark(corange))
    
    par(mar=c(2,2,2,2)+0.1)
    counts <- table(early$country)
    pie(counts, paste0(names(counts), " (", counts,")"), border="white", col=pal.dark(c(cgreen, corange, cblue)), cex=0.8)


    par(mar=c(5,4,0,0)+0.1)
    plotProportions(table(early$location), labels=levels(early$location), prop=FALSE, cex.axis=0.7, cex.names=0.7, cex.counts=0.6, col=pal.dark(cblue))

```


\clearpage


## Subsampled complete outbreak

Complete outbreak subsampled to approximately a quarter of sequences and roughly stratified by month.

```{r special-suball-date, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}

    set.seed(25)

    suball  <- seqdb[downsampleStratifiedByMonth(seqdb$date, 0.2),]
    write.csv(suball, paste0(outputpath,"special-suball.csv"), quote=FALSE, row.names=FALSE)
    
    plotSamplingHist(suball$date, daterange=range(seqdb$date), ymax=50, ylab="", xlab=paste0("Subsampled complete outbreak\n(",nrow(suball)," sequences)"), cex.years=1, cex.months=0.7, cex.axis=0.8, col=pal.dark(cblue))
    plotSamplingHist(suball$date, ymax=50, ylab="", xlab=paste0("Subsampled complete outbreak\n(",nrow(suball)," sequences)"), cex.years=1, cex.months=0.7, cex.axis=0.8, breaks='months', col=pal.dark(cgreen))

    stats <- addStats(stats, suball, "special-suball")


```

\clearpage

```{r special-suball-geo, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=3.5, fig.height=3}
    plotMapGradient(getDistrictCounts(suball[which(suball$country == "GIN"),]), GIN_adm2, numlabels=5, lwd=0.5, col=pal.dark(cgreen))
    plotMapGradient(getDistrictCounts(suball[which(suball$country == "SLE"),]), SLE_adm2, numlabels=5, lwd=0.5, col=pal.dark(cblue))
    plotMapGradient(getDistrictCounts(suball[which(suball$country == "LBR"),]), LBR_adm1, numlabels=5, lwd=0.5, col=pal.dark(corange))
    
    par(mar=c(2,2,2,2)+0.1)
    counts <- table(suball$country)
    pie(counts, paste0(names(counts), " (", counts,")"), border="white", col=pal.dark(c(cgreen, corange, cblue)), cex=0.8)

```

\clearpage

## Subsampled complete outbreak big

Complete outbreak subsampled to approximately half of all sequences and roughly stratified by month.

```{r special-subbig-date, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}

    set.seed(25)

    subbig  <- seqdb[downsampleStratifiedByMonth(seqdb$date, 0.5),]
    write.csv(subbig, paste0(outputpath,"special-subbig.csv"), quote=FALSE, row.names=FALSE)
    
    plotSamplingHist(subbig$date, daterange=range(seqdb$date), ymax=50, ylab="", xlab=paste0("Subsampled complete outbreak\n(",nrow(subbig)," sequences)"), cex.years=1, cex.months=0.7, cex.axis=0.8, col=pal.dark(cblue))
    plotSamplingHist(subbig$date, ymax=120, ylab="", xlab=paste0("Subsampled complete outbreak\n(",nrow(subbig)," sequences)"), cex.years=1, cex.months=0.7, cex.axis=0.8, breaks='months', col=pal.dark(cgreen))

    stats <- addStats(stats, subbig, "special-subbig")


```

\clearpage

```{r special-subbig-geo, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=3.5, fig.height=3}
    plotMapGradient(getDistrictCounts(suball[which(subbig$country == "GIN"),]), GIN_adm2, numlabels=5, lwd=0.5, col=pal.dark(cgreen))
    plotMapGradient(getDistrictCounts(suball[which(subbig$country == "SLE"),]), SLE_adm2, numlabels=5, lwd=0.5, col=pal.dark(cblue))
    plotMapGradient(getDistrictCounts(suball[which(subbig$country == "LBR"),]), LBR_adm1, numlabels=5, lwd=0.5, col=pal.dark(corange))
    
    par(mar=c(2,2,2,2)+0.1)
    counts <- table(subbig$country)
    pie(counts, paste0(names(counts), " (", counts,")"), border="white", col=pal.dark(c(cgreen, corange, cblue)), cex=0.8)

```

---


# Dataset statistics

```{r echo=FALSE, results='asis'}

    stats <- data.frame(stats)
    stats$earliest <- as.Date(stats$earliest, origin=dateorigin)
    stats$latest <- as.Date(stats$latest, origin=dateorigin)
    
    kable(stats,  caption="Dataset statistics")
```


# Select sequences from Fasta file

```{bash, results='hide', eval=TRUE}
    
    # Load Conda environment
    source /Users/user/anaconda3/envs/bio/bin/activate bio

    # Extract sequences
    for i in `ls ../results/datasets/*.csv` 
    do 
        # Full-length genomes
        python ../scripts/selectsequences.py -i ${i} -a ../data/sequence/Data/Makona_1610_genomes_2016-06-23.fasta -o ../results/datasets/nflg/ -m -u
        
        # Coding sequences
        python ../scripts/selectsequences.py -i ${i} -a ../results/datasets/Makona_1610_cds.trimmed.fas -o ../results/datasets/cds/ -m -u
        
        # Noncoding sequences
        python ../scripts/selectsequences.py -i ${i} -a ../results/datasets/Makona_1610_ig.trimmed.fas -o ../results/datasets/ig/ -m -u
    done
    
```

