---
title: "EBOV empirical sampling proportions"
author: "Louis du Plessis"
date: \today
fontsize: 10pt
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    keep_tex: false
    fig_crop: false
---

# Summary

Extract the weekly and monthly empirical sampling proportions for the datasets extracted in `extract_datasets.rmd` using surveillance data from the epidemic.
The empirical sampling proportion for a time period is defined as the number of sequences generated divided by the number of cases reported.


```{r rsetup, echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.path="figs/", fig.width=8, fig.height=4}

    rm(list = ls())
    library(gplots)
    library(bdskytools)
    #library(knitr)
    
    source("../scripts/PlotUtils.R")

  
    survpath  <- "../data/cases/"
    seqpath   <- "../results/datasets/"
    
    outputpath <- "../results/datasets/proportions/"    
    dir.create(outputpath, recursive=TRUE, showWarnings=FALSE)

    
    dateorigin <- "1970-01-01"

    parseMonth <- function(d) {
      return (format.Date(as.Date(d), "%Y-%m-01"))
    }
    
    sumMonths <- function(cases) {
      
      month <- parseMonth(rownames(cases)[1])
      monthcases <- c()
      monthsums  <- c()
      monthdates <- c(month)
      for (i in 1:nrow(cases)) {
          m <- parseMonth(rownames(cases)[i])
          if (m != month) {
              monthsums <- rbind(monthsums, colSums(monthcases))
              month     <- m
              monthdates <- c(monthdates, month)
              monthcases <- c(cases[i,])
          } else {
              monthcases <- rbind(monthcases, cases[i,])
          }
      }
      monthsums <- rbind(monthsums, colSums(monthcases))
      rownames(monthsums) <- monthdates
      
      monthsums <- data.frame(monthsums)
      monthsums$patientSum   <- cumsum(monthsums$patient)
      monthsums$situationSum <- cumsum(monthsums$situation)
      monthsums$combinedSum  <- cumsum(monthsums$combined)
      
      return(monthsums)
      
    }
    
    
    getWeeklyProportions <- function(cases, sequences, sanitise=TRUE) {
      
      seq_hist   <- hist(as.Date(sequences$date), breaks='weeks', start.on.monday=FALSE, plot=FALSE)
      hist_dates <- as.Date(seq_hist$breaks, origin=dateorigin)
      
      start <- which(rownames(cases) == hist_dates[1])
      end   <- which(rownames(cases) == hist_dates[length(hist_dates)-1])
      
      proportions <- seq_hist$counts/cases[start:end,c("patient","situation","combined")]
      proportions <- cbind(getYearDate(rownames(proportions)), proportions)
      colnames(proportions)[1] <- "yeardate"
      
      if (sanitise) {
          proportions$patient[which(proportions$patient > 1)] <- 1
          proportions$situation[which(proportions$situation > 1)] <- 1
          proportions$combined[which(proportions$combined > 1)] <- 1
          
          proportions$patient[which(is.nan(proportions$patient))] <- 0
          proportions$situation[which(is.nan(proportions$situation))] <- 0
          proportions$combined[which(is.nan(proportions$combined))] <- 0
      }
          
      return(proportions)
    }
    
    getMonthlyProportions <- function(cases, sequences, sanitise=TRUE) {
      
      seq_hist   <- hist(as.Date(sequences$date), breaks='months', right=FALSE, plot=FALSE)
      hist_dates <- as.Date(seq_hist$breaks, origin=dateorigin)
      
      start <- which(rownames(cases) == hist_dates[1])
      end   <- which(rownames(cases) == hist_dates[length(hist_dates)-1])
      
      proportions <- seq_hist$counts/cases[start:end,c("patient","situation","combined")]
      proportions <- cbind(getYearDate(rownames(proportions)), proportions)
      colnames(proportions)[1] <- "yeardate"
      
      if (sanitise) {
        proportions$patient[which(proportions$patient > 1)] <- 1
        proportions$situation[which(proportions$situation > 1)] <- 1
        proportions$combined[which(proportions$combined > 1)] <- 1
        
        proportions$patient[which(is.nan(proportions$patient))] <- 0
        proportions$situation[which(is.nan(proportions$situation))] <- 0
        proportions$combined[which(is.nan(proportions$combined))] <- 0
      }
      
      return(proportions)
    }
    
    # Assume cases are in order
    sumBins <- function(cases, breaks) {
      
      daterows <- getYearDate(rownames(cases))
      maxrow   <- which.max(daterows >= breaks[length(breaks)])-1
      
      cases <- as.matrix(cases)
      
      binnedcases <- matrix(0, ncol=ncol(cases), nrow=length(breaks))
      bin <- 1
      for (i in 1:maxrow) {
           date <- getYearDate(rownames(cases)[i])
           if (date >= breaks[bin]) {
              bin <- bin + 1
           }
           
           binnedcases[bin,] <- binnedcases[bin,] + cases[i,]
      }
      
      binnedcases <- data.frame(binnedcases)
      colnames(binnedcases) <- colnames(cases)
      rownames(binnedcases) <- getDayDate(c(min(daterows), breaks[1:(length(breaks)-1)]))
      binnedcases$patientSum   <- cumsum(binnedcases$patient)
      binnedcases$situationSum <- cumsum(binnedcases$situation)
      binnedcases$combinedSum  <- cumsum(binnedcases$combined)
      
      return(binnedcases)
    }
      
    
    getBinnedProportions <- function(cases, sequences, breaks, sanitise=TRUE) {
      
      # hist_dates <- c(as.Date("0000-01-01"), getDayDate(breaks))
      seq_hist   <- hist(as.Date(sequences$date), breaks=getDayDate(breaks), right=FALSE, plot=FALSE)

      proportions <- seq_hist$counts/cases[2:nrow(cases),c("patient","situation","combined")]
      proportions <- cbind(getYearDate(rownames(proportions)), proportions)
      #proportions <- cbind(proportions, cases$combined)
      #proportions <- cbind(proportions, seq_hist$count )
      colnames(proportions)[1] <- "yeardate"
      
      if (sanitise) {
          proportions$patient[which(proportions$patient > 1)] <- 1
          proportions$situation[which(proportions$situation > 1)] <- 1
          proportions$combined[which(proportions$combined > 1)] <- 1
          
          proportions$patient[which(is.nan(proportions$patient))] <- 0
          proportions$situation[which(is.nan(proportions$situation))] <- 0
          proportions$combined[which(is.nan(proportions$combined))] <- 0
      }
          
      return(proportions)
    }
    
    plotCasesAndProportion <- function(dataset, proportions, breaks='weeks', label="Weekly cases", plotcounts=FALSE, col=pal.dark(cblue)) {
      
        par(mar=c(5,4,4,4)+0.1)
    
        caseslist <- c()
        i <- 1
        for (i in 1:nrow(dataset)) {
            caseslist <- c(caseslist, rep(rownames(dataset)[i], dataset$combined[i]))
            i <- i + 1
        }
        caseslist <- as.Date(caseslist)
        
        dates <- as.Date(hist(range(as.Date(rownames(proportions))), breaks=breaks, plot=FALSE)$breaks, origin=dateorigin)
        
        
        plotSamplingHist(caseslist, ylab=label, xlab="", plotcounts=plotcounts, breaks=breaks, col=col, daterange=range(as.Date(rownames(proportions))))
        
        # May be a little buggy...
        par(new=TRUE)
        dates <- as.Date(hist(range(as.Date(rownames(proportions))), breaks=breaks, plot=FALSE)$breaks, origin=dateorigin)
        plot(dates, c(proportions$combined, proportions$combined[nrow(proportions)]), type='s', 
             col=pal.dark(cred), lwd=2, lty=2,
             ylim=c(0,1), xlim=range(dates), 
             xaxs='i',yaxs='i', axes=FALSE, xlab='', ylab='')
        axis(4, las=1, cex.axis=0.8)
        mtext("Sampling proportion", side=4, line=2)
      
    }
    
```

---    
    
\clearpage

# Datasets

## Complete dataset 

```{r all-cases, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}

    sle_cases <- read.table(paste0(survpath,"sierraleone_cases.csv"), header=TRUE)
    gin_cases <- read.table(paste0(survpath,"guinea_cases.csv"), header=TRUE)
    lbr_cases <- read.table(paste0(survpath,"liberia_cases.csv"), header=TRUE)
    all_cases <- sle_cases + gin_cases + lbr_cases
    
    all_sequences    <- read.csv(paste0(seqpath,"complete.csv"))
    all_weekly <- getWeeklyProportions(all_cases, all_sequences)
    write.csv(t(all_weekly), paste0(outputpath,"all-weekly.csv"), quote=FALSE)
    
    all_cases_monthly <- sumMonths(all_cases)
    all_monthly       <- getMonthlyProportions(all_cases_monthly, all_sequences)
    write.csv(t(all_monthly), paste0(outputpath,"all-monthly.csv"), quote=FALSE)

    daterange  <- range(getYearDate(all_sequences$date))
    datebreaks <- seq(daterange[1], daterange[2], length.out=10)
    all_cases_10bins <- sumBins(all_cases, breaks=datebreaks)
    all_10bins       <- getBinnedProportions(all_cases_10bins, all_sequences, breaks=datebreaks)
    write.csv(t(all_10bins), paste0(outputpath,"all-10bins.csv"), quote=FALSE)
    
    plotCasesAndProportion(all_cases, all_weekly,  breaks='weeks', label="Weekly cases",  plotcounts=FALSE, col=pal.dark(cblue))
    plotCasesAndProportion(all_cases, all_monthly, breaks='months',label="Monthly cases", plotcounts=FALSE, col=pal.dark(corange))
    
```



\clearpage

## Southeast Sierra Leone

```{r region-sle-southeast-cases, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}

    sle_southeast_cases     <- read.table(paste0(survpath,"sierraleone_southeast_cases.csv"), header=TRUE)
    sle_southeast_sequences <- read.csv(paste0(seqpath,"region-sle-southeast.csv"))
    sle_southeast_weekly    <- getWeeklyProportions(sle_southeast_cases, sle_southeast_sequences)
    write.csv(t(sle_southeast_weekly), paste0(outputpath,"sle-southeast-weekly.csv"), quote=FALSE)
    
    sle_southeast_cases_monthly <- sumMonths(sle_southeast_cases)
    sle_southeast_monthly       <- getMonthlyProportions(sle_southeast_cases_monthly, sle_southeast_sequences)
    write.csv(t(sle_southeast_monthly), paste0(outputpath,"sle-southeast-monthly.csv"), quote=FALSE)
        
    daterange  <- range(getYearDate(sle_southeast_sequences$date))
    datebreaks <- seq(daterange[1], daterange[2], length.out=10)
    sle_southeast_cases_10bins <- sumBins(sle_southeast_cases, breaks=datebreaks)
    sle_southeast_10bins       <- getBinnedProportions(sle_southeast_cases_10bins, sle_southeast_sequences, breaks=datebreaks)
    write.csv(t(sle_southeast_10bins), paste0(outputpath,"sle-southeast-10bins.csv"), quote=FALSE)
    
    
    plotCasesAndProportion(sle_southeast_cases, sle_southeast_weekly,  breaks='weeks', label="Weekly cases",  plotcounts=FALSE, col=pal.dark(cblue))
    plotCasesAndProportion(sle_southeast_cases, sle_southeast_monthly, breaks='months',label="Monthly cases", plotcounts=FALSE, col=pal.dark(corange))
    


```


## Western Sierra Leone

```{r region-sle-west-cases, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}

    sle_west_cases     <- read.table(paste0(survpath,"sierraleone_western_cases.csv"), header=TRUE)
    sle_west_sequences <- read.csv(paste0(seqpath,"region-sle-west.csv"))
    sle_west_weekly    <- getWeeklyProportions(sle_west_cases, sle_west_sequences)
    write.csv(t(sle_west_weekly), paste0(outputpath,"sle-west-weekly.csv"), quote=FALSE)
    
    sle_west_cases_monthly <- sumMonths(sle_west_cases)
    sle_west_monthly       <- getMonthlyProportions(sle_west_cases_monthly, sle_west_sequences)
    write.csv(t(sle_west_monthly), paste0(outputpath,"sle-west-monthly.csv"), quote=FALSE)
    
    daterange  <- range(getYearDate(sle_west_sequences$date))
    datebreaks <- seq(daterange[1], daterange[2], length.out=10)
    sle_west_cases_10bins <- sumBins(sle_west_cases, breaks=datebreaks)
    sle_west_10bins       <- getBinnedProportions(sle_west_cases_10bins, sle_west_sequences, breaks=datebreaks)
    write.csv(t(sle_west_10bins), paste0(outputpath,"sle-west-10bins.csv"), quote=FALSE)
    
    plotCasesAndProportion(sle_west_cases, sle_west_weekly,  breaks='weeks', label="Weekly cases",  plotcounts=FALSE, col=pal.dark(cblue))
    plotCasesAndProportion(sle_west_cases, sle_west_monthly, breaks='months',label="Monthly cases", plotcounts=FALSE, col=pal.dark(corange))
    
```


\clearpage


## Eastern Sierra Leone

```{r region-sle-east-cases, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}

    sle_east_cases     <- read.table(paste0(survpath,"sierraleone_eastern_cases.csv"), header=TRUE)
    sle_east_sequences <- read.csv(paste0(seqpath,"region-sle-east.csv"))
    sle_east_weekly    <- getWeeklyProportions(sle_east_cases, sle_east_sequences)
    write.csv(t(sle_east_weekly), paste0(outputpath,"sle-east-weekly.csv"), quote=FALSE)
    
    sle_east_cases_monthly <- sumMonths(sle_east_cases)
    sle_east_monthly       <- getMonthlyProportions(sle_east_cases_monthly, sle_east_sequences)
    write.csv(t(sle_east_monthly), paste0(outputpath,"sle-east-monthly.csv"), quote=FALSE)
    
    daterange  <- range(getYearDate(sle_east_sequences$date))
    datebreaks <- seq(daterange[1], daterange[2], length.out=10)
    sle_east_cases_10bins <- sumBins(sle_east_cases, breaks=datebreaks)
    sle_east_10bins       <- getBinnedProportions(sle_east_cases_10bins, sle_east_sequences, breaks=datebreaks)
    write.csv(t(sle_east_10bins), paste0(outputpath,"sle-east-10bins.csv"), quote=FALSE)
    
    plotCasesAndProportion(sle_east_cases, sle_east_weekly,  breaks='weeks', label="Weekly cases",  plotcounts=FALSE, col=pal.dark(cblue))
    plotCasesAndProportion(sle_east_cases, sle_east_monthly, breaks='months',label="Monthly cases", plotcounts=FALSE, col=pal.dark(corange))
    
    
```


\clearpage

## Liberia

```{r country-lbr-cases, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}

    lbr_cases     <- read.table(paste0(survpath,"liberia_cases.csv"), header=TRUE)
    lbr_sequences <- read.csv(paste0(seqpath,"country-lbr.csv"))
    lbr_weekly    <- getWeeklyProportions(lbr_cases, lbr_sequences)
    write.csv(t(lbr_weekly), paste0(outputpath,"lbr-weekly.csv"), quote=FALSE)
    
    lbr_cases_monthly <- sumMonths(lbr_cases)
    lbr_monthly       <- getMonthlyProportions(lbr_cases_monthly, lbr_sequences)
    write.csv(t(lbr_monthly), paste0(outputpath,"lbr-monthly.csv"), quote=FALSE)
    
    daterange  <- range(getYearDate(lbr_sequences$date))
    datebreaks <- seq(daterange[1], daterange[2], length.out=10)
    lbr_cases_10bins <- sumBins(lbr_cases, breaks=datebreaks)
    lbr_10bins       <- getBinnedProportions(lbr_cases_10bins, lbr_sequences, breaks=datebreaks)
    write.csv(t(lbr_10bins), paste0(outputpath,"lbr-10bins.csv"), quote=FALSE)
    
    plotCasesAndProportion(lbr_cases, lbr_weekly,  breaks='weeks', label="Weekly cases",  plotcounts=FALSE, col=pal.dark(cblue))
    plotCasesAndProportion(lbr_cases, lbr_monthly, breaks='months',label="Monthly cases", plotcounts=FALSE, col=pal.dark(corange))
    
    
```


\clearpage


## Early outbreak

All sequences until 4 August (WHO declares public health emergency). 

```{r special-early, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}

    sle_cases <- read.table(paste0(survpath,"sierraleone_cases.csv"), header=TRUE)
    gin_cases <- read.table(paste0(survpath,"guinea_cases.csv"), header=TRUE)
    lbr_cases <- read.table(paste0(survpath,"liberia_cases.csv"), header=TRUE)
    all_cases <- sle_cases + gin_cases + lbr_cases
    
    early_sequences    <- read.csv(paste0(seqpath,"special-early.csv"))
    early_weekly <- getWeeklyProportions(all_cases, early_sequences)
    write.csv(t(early_weekly), paste0(outputpath,"special-early-weekly.csv"), quote=FALSE)
    
    all_cases_monthly <- sumMonths(all_cases)
    early_monthly       <- getMonthlyProportions(all_cases_monthly, early_sequences)
    write.csv(t(early_monthly), paste0(outputpath,"special-early-monthly.csv"), quote=FALSE)
    
    daterange  <- range(getYearDate(early_sequences$date))
    datebreaks <- seq(daterange[1], daterange[2], length.out=10)
    all_cases_10bins <- sumBins(all_cases, breaks=datebreaks)
    early_10bins       <- getBinnedProportions(all_cases_10bins, early_sequences, breaks=datebreaks)
    write.csv(t(early_10bins), paste0(outputpath,"special-early-10bins.csv"), quote=FALSE)

    plotCasesAndProportion(all_cases, early_weekly,  breaks='weeks', label="Weekly cases",  plotcounts=FALSE, col=pal.dark(cblue))
    plotCasesAndProportion(all_cases, early_monthly, breaks='months',label="Monthly cases", plotcounts=FALSE, col=pal.dark(corange))
    
```




\clearpage


## Subsampled complete outbreak

Complete outbreak subsampled to approximately a quarter of sequences and roughly stratified by month.

```{r special-suball, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}

    sle_cases <- read.table(paste0(survpath,"sierraleone_cases.csv"), header=TRUE)
    gin_cases <- read.table(paste0(survpath,"guinea_cases.csv"), header=TRUE)
    lbr_cases <- read.table(paste0(survpath,"liberia_cases.csv"), header=TRUE)
    all_cases <- sle_cases + gin_cases + lbr_cases
    
    sub_sequences    <- read.csv(paste0(seqpath,"special-suball.csv"))
    sub_weekly <- getWeeklyProportions(all_cases, sub_sequences)
    write.csv(t(sub_weekly), paste0(outputpath,"special-suball-weekly.csv"), quote=FALSE)
    
    all_cases_monthly <- sumMonths(all_cases)
    sub_monthly       <- getMonthlyProportions(all_cases_monthly, sub_sequences)
    write.csv(t(sub_monthly), paste0(outputpath,"special-suball-monthly.csv"), quote=FALSE)
    
    daterange  <- range(getYearDate(sub_sequences$date))
    datebreaks <- seq(daterange[1], daterange[2], length.out=10)
    all_cases_10bins <- sumBins(all_cases, breaks=datebreaks)
    sub_10bins       <- getBinnedProportions(all_cases_10bins, sub_sequences, breaks=datebreaks)
    write.csv(t(sub_10bins), paste0(outputpath,"special-suball-10bins.csv"), quote=FALSE)

    plotCasesAndProportion(all_cases, sub_weekly,  breaks='weeks', label="Weekly cases",  plotcounts=FALSE, col=pal.dark(cblue))
    plotCasesAndProportion(all_cases, sub_monthly, breaks='months',label="Monthly cases", plotcounts=FALSE, col=pal.dark(corange))
    
```


\clearpage

## Subsampled complete outbreak big

Complete outbreak subsampled to approximately half of all sequences and roughly stratified by month.

```{r special-subbig, echo=FALSE, warning=FALSE, results='hide', fig.path=paste0(outputpath,"figs/"), fig.width=8, fig.height=4}

    sle_cases <- read.table(paste0(survpath,"sierraleone_cases.csv"), header=TRUE)
    gin_cases <- read.table(paste0(survpath,"guinea_cases.csv"), header=TRUE)
    lbr_cases <- read.table(paste0(survpath,"liberia_cases.csv"), header=TRUE)
    all_cases <- sle_cases + gin_cases + lbr_cases
    
    sub_sequences    <- read.csv(paste0(seqpath,"special-subbig.csv"))
    sub_weekly <- getWeeklyProportions(all_cases, sub_sequences)
    write.csv(t(sub_weekly), paste0(outputpath,"special-subbig-weekly.csv"), quote=FALSE)
    
    all_cases_monthly <- sumMonths(all_cases)
    sub_monthly       <- getMonthlyProportions(all_cases_monthly, sub_sequences)
    write.csv(t(sub_monthly), paste0(outputpath,"special-subbig-monthly.csv"), quote=FALSE)

    daterange  <- range(getYearDate(sub_sequences$date))
    datebreaks <- seq(daterange[1], daterange[2], length.out=10)
    all_cases_10bins <- sumBins(all_cases, breaks=datebreaks)
    sub_10bins       <- getBinnedProportions(all_cases_10bins, sub_sequences, breaks=datebreaks)
    write.csv(t(sub_10bins), paste0(outputpath,"special-subbig-10bins.csv"), quote=FALSE)
    
    plotCasesAndProportion(all_cases, sub_weekly,  breaks='weeks', label="Weekly cases",  plotcounts=FALSE, col=pal.dark(cblue))
    plotCasesAndProportion(all_cases, sub_monthly, breaks='months',label="Monthly cases", plotcounts=FALSE, col=pal.dark(corange))
    
```


